---
title: "STAT406 - Lecture 16 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

Preliminary lecture slides will be here.


# Random Forests (continued)

Consider again the ISOLET data, available 
here: 
[http://archive.ics.uci.edu/ml/datasets/ISOLET](http://archive.ics.uci.edu/ml/datasets/ISOLET). 
Here we only use a subset 
corresponding to the observations for the letters **C** and **Z**. 

We first load the training and test data sets, and force the response 
variable to be categorical, so that the `R` implementations of the
different predictors we will use below will build 
classifiers and not their regression counterparts:
```{r rf.isolet, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
xtr <- read.table('../Lecture15/isolet-train-c-z.data', sep=',')
xte <- read.table('../Lecture15/isolet-test-c-z.data', sep=',') 
xtr$V618 <- as.factor(xtr$V618)
xte$V618 <- as.factor(xte$V618)
```
To train a Random Forest we use the function `randomForest` in the
package of the same name. The code underlying this package was originally 
written by Leo Breiman. We train a RF leaving all
paramaters at their default values, and check 
its performance on the test set:
```{r rf.isolet2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(randomForest)
set.seed(123)
a.rf <- randomForest(V618 ~ ., data=xtr, ntree=500) 
p.rf <- predict(a.rf, newdata=xte, type='response')
table(p.rf, xte$V618)
```
Note that the Random Forest only makes one mistake out of 120 observations
in the test set. The OOB error rate estimate is slightly over 2%, 
and we see that 500 trees is a reasonable forest size:

```{r rf.oob, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(a.rf, lwd=3, lty=1)
a.rf
```

#### Using a test set instead of OBB

Given that in this case we do have a test set, we can use it 
to monitor the error rate (instead of using the OOB error estimates):

```{r rf.isolet.test, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
x.train <- model.matrix(V618 ~ ., data=xtr)
y.train <- xtr$V618
x.test <- model.matrix(V618 ~ ., data=xte)
y.test <- xte$V618
set.seed(123)
a.rf <- randomForest(x=x.train, y=y.train, xtest=x.test, ytest=y.test, ntree=500) 
test.err <- a.rf$test$err.rate
ma <- max(c(test.err))
plot(test.err[, 2], lwd=2, lty=1, col='red', type='l', ylim=c(0, max(c(0, ma))))
lines(test.err[, 3], lwd=2, lty=1, col='green')
lines(test.err[, 1], lwd=2, lty=1, col='black')
```

According to the help page for the `plot` method for objects of class
`randomForest`, the following plot should show both error rates (OOB plus
those on the test set):

```{r rf.isolet.test.plot, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(a.rf, lwd=2)
a.rf
```

#### Feature sequencing / Variable ranking

To explore which variables were used in the forest,
and also, their importance rank as discussed in
class, we can use the function `varImpPlot`:
```{r rf.isolet3, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
varImpPlot(a.rf, n.var=20)
```

#### Comparing RF with other classifiers

We now compare the Random Forest with some of the other classifiers we saw in class,
using their classification error rate on the test set as our comparison measure. 
We first start with K-NN:
```{r rf.isolet4, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(class)
u1 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 1)
table(u1, xte$V618)

u5 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 5)
table(u5, xte$V618)

u10 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 10)
table(u10, xte$V618)

u20 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 20)
table(u20, xte$V618)

u50 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 50)
table(u50, xte$V618)
```
To use logistic regression we first create a new variable that is 1
for the letter **C** and 0 for the letter **Z**, and use it as
our response variable. 
```{r rf.isoletglm, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
xtr$V619 <- as.numeric(xtr$V618==3)
d.glm <- glm(V619 ~ . - V618, data=xtr, family=binomial)
pr.glm <- as.numeric( predict(d.glm, newdata=xte, type='response') >  0.5 )
table(pr.glm, xte$V618)
```
Question for the reader: why do you think this classifier's performance
is so disappointing? 

It is interesting to see how a simple LDA classifier does:
```{r rf.isolet5, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(MASS)
xtr$V619 <- NULL
d.lda <- lda(V618 ~ . , data=xtr)
pr.lda <- predict(d.lda, newdata=xte)$class
table(pr.lda, xte$V618)
```
Finally, note that a carefully built classification tree 
performs remarkably well, only using 3 features:
```{r rf.isolet6, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(rpart)
my.c <- rpart.control(minsplit=5, cp=1e-8, xval=10)
set.seed(987)
a.tree <- rpart(V618 ~ ., data=xtr, method='class', parms=list(split='information'), control=my.c)
cp <- a.tree$cptable[which.min(a.tree$cptable[,"xerror"]),"CP"]
a.tp <- prune(a.tree, cp=cp)
p.t <- predict(a.tp, newdata=xte, type='vector')
table(p.t, xte$V618)
```
Finally, note that if you train a single classification tree with the
default values for the stopping criterion tuning parameters, the 
tree also uses only 3 features, but its classification error rate
on the test set is larger than that of the pruned one:
```{r rf.isolet7, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
set.seed(987)
a2.tree <- rpart(V618 ~ ., data=xtr, method='class', parms=list(split='information'))
p2.t <- predict(a2.tree, newdata=xte, type='vector')
table(p2.t, xte$V618)
```

