---
title: "STAT406 - Lecture 16 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=6, message=FALSE, warning=FALSE)
```

## LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

Preliminary lecture slides will be here.

# Random Forests

Even though using a *bagged* ensemble of trees usually results in a more
stable predictor / classifier, a better ensemble can be improved by training 
each of its members in a careful way. The main idea is to try to reduce the 
(conditional) potential correlation among the predictions of the bagged trees, 
as discussed in class. Each of the bootstrap trees in the ensemble is
grown using only a randomly selected set of features when partitioning each node.
More specifically, at each node only a random subset of explanatory variables
is considered to determine the optimal split. This randomly chosen features 
are selected independently at each node as the tree is being constructed. 

To train a Random Forest in `R` we use the 
funtion `randomForest` from the package with the same name.
The syntax is the same as that of `rpart`, but the tuning parameters 
for each of the *trees* in the *forest* are different from `rpart`. 
Refer to the help page if you need to modify them. 

We load and prepare the admissions data as before:
```{r init}
mm <- read.table('../Lecture14/T11-6.DAT', header=FALSE)
mm$V3 <- as.factor(mm$V3)
mm[,2] <- mm[,2] / 150
```
and train a Random Forest with 500 trees and using all the default tuning parameters:
```{r rf1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(randomForest)
a.rf <- randomForest(V3~V1+V2, data=mm, ntree=500) 
```
Predictions can be obtained using the `predict` method, as usual, when 
you specify the `newdata` argument. Refer to the help page 
of `predict.randomForest` for details on the different 
behaviour of `predict` for Random Forests objects when the argument `newdata` is 
either present or missing. 

To visualize the predicted classes obtained with a Random Forest
on our example data, we compute the corresponding
predicted conditional class probabilities on the 
same grid used before:
```{r inst2.0}
aa <- seq(2, 5, length=200)
bb <- seq(2, 5, length=200)
dd <- expand.grid(aa, bb)
names(dd) <- names(mm)[1:2]
```
The estimated conditional probabilities for class *red* 
are shown in the plot below
(how are these estimated conditional probabilities computed exactly?)
```{r rf1.1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
pp.rf <- predict(a.rf, newdata=dd, type='prob')
filled.contour(aa, bb, matrix(pp.rf[,1], 200, 200), col=terrain.colors(20), xlab='GPA', ylab='GMAT', 
               plot.axes={axis(1); axis(2)},
               panel.last={points(mm[,-3], pch=19, cex=1.5, col=c("red", "blue", "green")[mm[,3]])}
               )
```

And the predicted conditional probabilities for the rest of the classes are:
```{r rf2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
filled.contour(aa, bb, matrix(pp.rf[,2], 200, 200), col=terrain.colors(20), xlab='GPA', 
               ylab='GMAT', plot.axes={axis(1); axis(2)}, 
               panel.last={points(mm[,-3], pch=19, cex=1.5, col=c("red", "blue", "green")[mm[,3]]) }
               )
filled.contour(aa, bb, matrix(pp.rf[,3], 200, 200), col=terrain.colors(20), xlab='GPA', 
               ylab='GMAT', plot.axes={axis(1); axis(2)}, 
               panel.last={points(mm[,-3], pch=19, cex=1.5, col=c("red", "blue", "green")[mm[,3]]) }
               )
```

A very interesting exercise would be to train a Random Forest on the perturbed data
(in `mm2`) and verify that the predicted conditional probabilities do not change much, 
as was the case for the bagged classifier. 

### Another example

We will now use a more interesting example. The ISOLET data, available 
here: 
[http://archive.ics.uci.edu/ml/datasets/ISOLET](http://archive.ics.uci.edu/ml/datasets/ISOLET), 
contains data 
on sound recordings of 150 speakers saying each letter of the
alphabet (twice). See the original source for more details. Since 
the full data set is rather large, here we only use the subset 
corresponding to the observations for the letters **C** and **Z**. 

We first load the training and test data sets, and force the response 
variable to be categorical, so that the `R` implementations of the
different predictors we will use below will build 
classifiers and not their regression counterparts:
```{r rf.isolet, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
xtr <- read.table('../Lecture15/isolet-train-c-z.data', sep=',')
xte <- read.table('../Lecture15/isolet-test-c-z.data', sep=',') 
xtr$V618 <- as.factor(xtr$V618)
xte$V618 <- as.factor(xte$V618)
```
To train a Random Forest we use the function `randomForest` in the
package of the same name. The code underlying this package was originally 
written by Leo Breiman. We first train a Random Forest, using all the default parameters
```{r rf.isolet02, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(randomForest)
set.seed(123)
( a.rf <- randomForest(V618 ~ ., data=xtr, ntree=500) )
```
We now check its performance on the test set:
```{r rf.isolet.te}
p.rf <- predict(a.rf, newdata=xte, type='response')
table(p.rf, xte$V618)
```
Note that the Random Forest only makes one mistake out of 120 (approx 0.8%) observations
in the test set. However, the OOB error rate estimate is slightly over 2%. 
The next plot shows the evolution of the OOB error rate estimate as a function of the 
number of classifiers in the ensemble (trees in the forest). Note that 500 trees 
appears to be a reasonable forest size, in the sense thate the OOB error rate estimate is stable. 
```{r rf0.oob, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(a.rf, lwd=3, lty=1)
```

<!-- Consider again the ISOLET data, available  -->
<!-- here:  -->
<!-- [http://archive.ics.uci.edu/ml/datasets/ISOLET](http://archive.ics.uci.edu/ml/datasets/ISOLET).  -->
<!-- Here we only use a subset  -->
<!-- corresponding to the observations for the letters **C** and **Z**.  -->

<!-- We first load the training and test data sets, and force the response  -->
<!-- variable to be categorical, so that the `R` implementations of the -->
<!-- different predictors we will use below will build  -->
<!-- classifiers and not their regression counterparts: -->
<!-- ```{r rf.isolet, fig.width=6, fig.height=6, message=FALSE, warning=FALSE} -->
<!-- xtr <- read.table('../Lecture15/isolet-train-c-z.data', sep=',') -->
<!-- xte <- read.table('../Lecture15/isolet-test-c-z.data', sep=',')  -->
<!-- xtr$V618 <- as.factor(xtr$V618) -->
<!-- xte$V618 <- as.factor(xte$V618) -->
<!-- ``` -->

<!-- To train a Random Forest we use the function `randomForest` in the -->
<!-- package of the same name. The code underlying this package was originally  -->
<!-- written by Leo Breiman. We train a RF leaving all -->
<!-- paramaters at their default values, and check  -->
<!-- its performance on the test set: -->
<!-- ```{r rf.isolet2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE} -->
<!-- library(randomForest) -->
<!-- set.seed(123) -->
<!-- a.rf <- randomForest(V618 ~ ., data=xtr, ntree=500)  -->
<!-- p.rf <- predict(a.rf, newdata=xte, type='response') -->
<!-- table(p.rf, xte$V618) -->
<!-- ``` -->
<!-- Note that the Random Forest only makes one mistake out of 120 observations -->
<!-- in the test set. The OOB error rate estimate is slightly over 2%,  -->
<!-- and we see that 500 trees is a reasonable forest size: -->

<!-- ```{r rf.oob, fig.width=6, fig.height=6, message=FALSE, warning=FALSE} -->
<!-- plot(a.rf, lwd=3, lty=1) -->
<!-- a.rf -->
<!-- ``` -->

#### Using a test set instead of OBB

Given that in this case we do have a test set, we can use it 
to monitor the error rate (instead of using the OOB error estimates):

```{r rf.isolet.test, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
x.train <- model.matrix(V618 ~ ., data=xtr)
y.train <- xtr$V618
x.test <- model.matrix(V618 ~ ., data=xte)
y.test <- xte$V618
set.seed(123)
a.rf <- randomForest(x=x.train, y=y.train, xtest=x.test, ytest=y.test, ntree=500) 
test.err <- a.rf$test$err.rate
ma <- max(c(test.err))
plot(test.err[, 2], lwd=2, lty=1, col='red', type='l', ylim=c(0, max(c(0, ma))))
lines(test.err[, 3], lwd=2, lty=1, col='green')
lines(test.err[, 1], lwd=2, lty=1, col='black')
```

According to the help page for the `plot` method for objects of class
`randomForest`, the following plot should show both error rates (OOB plus
those on the test set):

```{r rf.isolet.test.plot, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(a.rf, lwd=2)
```

#### Feature sequencing / Variable ranking

To explore which variables were used in the forest,
and also, their importance rank as discussed in
class, we can use the function `varImpPlot`:
```{r rf.isolet3, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
varImpPlot(a.rf, n.var=20)
```

#### Comparing RF with other classifiers

We now compare the Random Forest with some of the other classifiers we saw in class,
using their classification error rate on the test set as our comparison measure. 
We first start with K-NN:
```{r rf.isolet4, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(class)
u1 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 1)
table(u1, xte$V618)

u5 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 5)
table(u5, xte$V618)

u10 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 10)
table(u10, xte$V618)

u20 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 20)
table(u20, xte$V618)

u50 <- knn(train=xtr[, -618], test=xte[, -618], cl=xtr[, 618], k = 50)
table(u50, xte$V618)
```
To use logistic regression we first create a new variable that is 1
for the letter **C** and 0 for the letter **Z**, and use it as
our response variable. 
```{r rf.isoletglm, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
xtr$V619 <- as.numeric(xtr$V618==3)
d.glm <- glm(V619 ~ . - V618, data=xtr, family=binomial)
pr.glm <- as.numeric( predict(d.glm, newdata=xte, type='response') >  0.5 )
table(pr.glm, xte$V618)
```
Question for the reader: why do you think this classifier's performance
is so disappointing? 

It is interesting to see how a simple LDA classifier does:
```{r rf.isolet5, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(MASS)
xtr$V619 <- NULL
d.lda <- lda(V618 ~ . , data=xtr)
pr.lda <- predict(d.lda, newdata=xte)$class
table(pr.lda, xte$V618)
```
Finally, note that a carefully built classification tree 
performs remarkably well, only using 3 features:
```{r rf.isolet6, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(rpart)
my.c <- rpart.control(minsplit=5, cp=1e-8, xval=10)
set.seed(987)
a.tree <- rpart(V618 ~ ., data=xtr, method='class', parms=list(split='information'), control=my.c)
cp <- a.tree$cptable[which.min(a.tree$cptable[,"xerror"]),"CP"]
a.tp <- prune(a.tree, cp=cp)
p.t <- predict(a.tp, newdata=xte, type='vector')
table(p.t, xte$V618)
```
Finally, note that if you train a single classification tree with the
default values for the stopping criterion tuning parameters, the 
tree also uses only 3 features, but its classification error rate
on the test set is larger than that of the pruned one:
```{r rf.isolet7, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
set.seed(987)
a2.tree <- rpart(V618 ~ ., data=xtr, method='class', parms=list(split='information'))
p2.t <- predict(a2.tree, newdata=xte, type='vector')
table(p2.t, xte$V618)
```

