---
title: "STAT406 - Lecture 12 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

Preliminary lecture slides will be here.


## Bagging

One strategy to obtain more stable predictors is called
**Bootstrap AGGregatING** (bagging). It can be applied to
many predictors (not only trees), and it generally results
in larger improvements in prediction quality when it is used with predictors
that are flexible (low bias), but highly variable.

The justification and motivation were discussed in class. Intuitively
we are averaging the predictions obtained from an estimate of the 
"average prediction" we would have computed had we had access to 
several (many?) independent training sets (samples). 

There are several (many?) `R` packages implementing
bagging for different predictors, with varying degrees of 
flexibility (the implementations) and user-friendliness. 
However, for pedagogical and illustrative purposes, in these notes I will
*bagg* by hand.

### Bagging by hand

Again, to simplify the discussion and presentation, in order to evaluate 
prediction quality I will split the 
data (`Boston`) into a training and a test set. We do this now:
```{r bag1, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
library(rpart)
data(Boston, package='MASS')
set.seed(123)
n <- nrow(Boston)
ii <- sample(n, floor(n/4))
dat.te <- Boston[ ii, ]
dat.tr <- Boston[ -ii, ]
```
I will now train $N = 5$ trees and average their predictions. 
Note that, in order to illustrate the process more
clearly, I will compute and store the $N \times n_e$
predictions, where $n_e$ denotes the number of observations in 
the test set. This is not the best (most efficient) way of implementing *bagging*,
but the main purpose here is to understand **what** we are doing. Also note that
an alternative (better in terms of reusability of the
ensemble, but maybe still not the most efficient option) would be
to store the $N$ trees directly. This would also allow for
more elegant and easy to read code. Once again, this approach 
will be sacrificed in the altar of clarity of presentation and 
pedagogy (but I do illustrate it below!)

First create an array where we will store all the predictions:
```{r bag2, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
N <- 5
myps <- array(NA, dim=c(nrow(dat.te), N))
con <- rpart.control(minsplit=3, cp=1e-3, xval=1)
```
The last object (`con`) contains my options to train large
(potentially overfitting) trees. 
```{r bag3, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
n.tr <- nrow(dat.tr)
set.seed(123)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```
And compare with predictions from the pruned tree, and the
ones from other predictors discussed in the previous note:
```{r bag4, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
myc <- rpart.control(minsplit=3, cp=1e-8, xval=10)
set.seed(123)
bos.to <- rpart(medv ~ ., data=dat.tr, method='anova',
                control=myc)
b <- bos.to$cptable[which.min(bos.to$cptable[,"xerror"]),"CP"]
bos.t3 <- prune(bos.to, cp=b)
pr.t3 <- predict(bos.t3, newdata=dat.te, type='vector')
with(dat.te, mean((medv - pr.t3)^2) )
```

What if we *bagg* $N = 10$ trees? 
```{r bag10, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
N <- 10
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```
or $N = 100$ trees? 
```{r bag100, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
N <- 100
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```
or $N = 1000$ trees? 
```{r bag1000, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
N <- 1000
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```

Note that, at least for this test set, increasing the number of bagged trees
seems to improve the MSPE. However, the gain appears to decrease, so it may
not be worth the computational effort to use a larger *bag* / ensemble. 
Furthermore, one may also want to investigate whether this is an
artifact of this specific training / test partition, or if similar
patterns of MSPE are observed for other random training / test splits. 
Below we try a different test/training split and repeat the 
bagging experiment above: 
```{r anothersplit, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
set.seed(123)
n <- nrow(Boston)
ii <- sample(n, floor(n/4))
dat.te <- Boston[ ii, ]
dat.tr <- Boston[ -ii, ]
for(N in c(5, 10, 100, 1000)) {
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
print(c(N, with(dat.te, mean( (medv - pr.bagg)^2 ) )))
}
```
The pattern is in fact similar to the one we observed before: 
increasing the size of the ensemble $N$ helps, but the improvement 
becomes smaller as $N$ increases. A very good exercise is to explore what happens with the MSPE of
the bagged ensemble when the MSPE is estimated using cross-validation
(instead of using a test set). I leave this as an exercise for the reader. 


#### More efficient, useful and elegant implementation  

I will now illustrate a possibly more efficient way to implement bagging, namely
storing the $N$ trees (rather than their predictions on a given data set).
In this way one can re-use the ensemble (on any future data set) without
having to re-train the elements of the *bag*. Since the idea is 
the same, I will just do it for ensemble of $N = 100$ trees. 
To simplify the comparison between this implementation of
bagging and the one used above, we first re-create
the original training / test split
```{r bag1000.alt0, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
set.seed(123)
n <- nrow(Boston)
ii <- sample(n, floor(n/4))
dat.te <- Boston[ ii, ]
dat.tr <- Boston[ -ii, ]
```
Now, let's create a `list` of 100 (empty) elements, each element of this 
list will store a regression tree:
```{r bag1000.alt, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
N <- 100
mybag <- vector('list', N)
```
Now, we train the $N$ trees as before, but store them in the `list` (without
computing any predictions):
```{r bag1000.alt2, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
set.seed(123)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  mybag[[j]] <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
}
```
Given a new data set, in order to obtain the corresponding predictions for 
each tree in the ensemble, one could either:

* loop over the $N$ trees, averaging the corresponding $N$ vectors of predictions; or
* use `sapply` (check the help page if you are not familiar with the `apply` functions in `R`).

The later option results in code that is much more elegant, 
efficient (allowing for future uses of the ensemble),
and compact. Of course both give exactly the same results. Below 
we illustrate both strategies. If we use the **first approach**
we obtain the following estimated MSPE using the test set:
```{r bag1000.alt3, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
pr.bagg2 <- rep(0, nrow(dat.te))
for(j in 1:N)
  pr.bagg2 <- pr.bagg2 + predict(mybag[[j]], newdata=dat.te) / N
with(dat.te, mean( (medv - pr.bagg2)^2 ) )
```
(compare it with the results we obtained before). Using the **second approach**:
```{r bag1000.alt4, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
pr.bagg3 <- rowMeans(sapply(mybag, predict, newdata=dat.te))
with(dat.te, mean( (medv - pr.bagg3)^2 ) )
```
Both results are of course identical. 


### Bagging a regression spline

Bagging does not provide much of an advantage when applied to linear
predictors (can you explain why?) Nevertheless, let us try it on the `lidar` data, 
which, as we did before, we randomly split into a training and test set:
```{r bagsplines, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
data(lidar, package='SemiPar')
set.seed(123)
n <- nrow(lidar)
ii <- sample(n, floor(n/5))
lid.te <- lidar[ ii, ]
lid.tr <- lidar[ -ii, ]
```
Now fit a cubic spline, and estimate the MSPE using the test set:
```{r bagsplines2, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
library(splines)
a <- lm(logratio ~ bs(x=range, df=10, degree=3), data=lid.tr) 
oo <- order(lid.tr$range)
pr.of <- predict(a, newdata=lid.te)
mean( (lid.te$logratio - pr.of)^2 )
```
We build an ensemble of 10 fits and estimate the corresponding
MSPE using the test set:
```{r bagsplines3, fig.width=5, fig.height=5, message=FALSE, warning=FALSE}
N <- 10 # 5 500 1500
myps <- matrix(NA, nrow(lid.te), N)
set.seed(123)
n.tr <- nrow(lid.tr)
for(i in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  a.b <- lm(logratio ~ bs(x=range, df=10, degree=3), data=lid.tr[ii,]) 
  myps[,i] <- predict(a.b, newdata=lid.te)
}
pr.ba <- rowMeans(myps)# , na.rm=TRUE)
mean( (lid.te$logratio - pr.ba)^2 )
```


