---
title: "STAT406 - Lecture 1 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides 

The lecture slides are [here](STAT406-18-lecture-1.pdf). 

## Predictions using a linear model
In this document we will explore (rather superficially)
some challenges found when trying to estimate the forecasting 
properties (e.g. the mean squared prediction error) of a (linear) predictor. We will
use the air-pollution data set, which I have split into a *training set* and a *test set*. 
The test set will be ignored when **training** our model (in the case of a linear
model, "**training**" simply means "**when estimating the vector of linear regression 
parameters**").

If you are interested in how these sets (*training* and *test*) were constructed: 
I ran the following script (you
do not need to do this, as I am providing both data sets to you, 
but you can re-create them yourself if you want to):
```{r construct, fig.width=5, fig.height=5, echo=TRUE}
x <- read.csv('rutgers-lib-30861_CSV-1.csv')
set.seed(123)
ii <- sample(rep(1:4, each=15))
# this is the training set `pollution-train.dat`
x.tr <- x[ii != 2, ]
# this is the test set `pollution-test.dat`
x.te <- x[ii == 2, ]
# then I saved them to disk:
# write.csv(x.tr, file='pollution-train.dat', row.names=FALSE, quote=FALSE)
# write.csv(x.te, file='pollution-test.dat', row.names=FALSE, quote=FALSE)
```

We now read the **training** data set from the file `pollution-train.dat`,
which is available [here](pollution-train.dat), and check that it was read properly:
```{r readtrain}
x.tr <- read.table('pollution-train.dat', header=TRUE, sep=',')
# sanity check
head(x.tr)
```
The response variable is `MORT`. 
Our first step is to fit a 
linear regression model with all available
predictors and look at a few diagnostic plots
where everything looks fine:
```{r full, fig.width=5, fig.height=5, echo=TRUE}
full <- lm(MORT ~ . , data=x.tr)
plot(full, which=1)
plot(full, which=2)
```

We also take a look at the estimated coeficients: 
```{r diag, fig.width=5, fig.height=5, echo=TRUE}
summary(full)
```

The fit appears to be routine, and reasonable (why? what did I check to come to this conclusion?). 

### A new focus: prediction
This course will be primarily concerned with making (good) predictions for cases 
(data points) that we may have not observed yet (future predictions). This is a bit
different from the focus of other Statistics courses you may have taken. You will
see later in the course that what you learned in other Statistics courses 
(e.g. trade-offs between flexibility and stability of different models, uncertainty
and standard techniques to reduce it, etc.) will prove
to be critical for building good predictions. 

As a simple example, in the rest of this note we will compare the quality of this model's predictions with those of a simpler (smaller) linear model with only 5 predictors. For this illustrative example, we will not worry about how these 5 explanatory variables were selected, however, this will play a **critical** role later in the course).

We now fit this **reduced** model and look at the estimated parameters and diagnostic plots
```{r reduced, fig.width=5, fig.height=5, echo=TRUE}
reduced <- lm(MORT ~ POOR + HC + NOX + HOUS + NONW, data=x.tr)
summary(reduced)
plot(reduced, which=1)
plot(reduced, which=2)
```

Although the reduced linear model (with 5 predictors) does not seem to provide a fit
as good as the one we get with full model, it is still acceptable. 
```{r gofs}
sum( resid(reduced)^2 )
sum( resid(full)^2 )
```
This observation should be obvious to you, since, 
as you already now, a model will **always** yield
a better fit to the data in terms of 
residual sum of squares than any of its submodels 
(i.e. any model using a subset of the explanatory
variables). I expect you to be able to formally 
prove the last satement. 

Our question of interest here is:
"Which model produces better predictions?" In many cases one is 
interested in predicting future observations, i.e. 
predicting the response variable for data
that was not available when the model / predictor was 
*fit* or *trained*. As we discussed in class, a reasonably
fair comparison can be obtined by 
comparing the mean squared predictions
of these two linear models on the test set, which we
read into `R` as follows:
```{r pred1}
x.te <- read.table('pollution-test.dat', header=TRUE, sep=',')
head(x.te)
```
Now compute the predicted values for the test set
with both the **full** and **reduced** models:
```{r pred2}
x.te$pr.full <- predict(full, newdata=x.te)  
x.te$pr.reduced <- predict(reduced, newdata=x.te)  
```
and compute the corresponding mean squared prediction errors:
```{r mspe}
with(x.te, mean( (MORT - pr.full)^2 ))
with(x.te, mean( (MORT - pr.reduced)^2 ))
```

Note that the reduced model (that did not fit the data
as well as the full model) nevertheless produced
better predictions (smaller mean squared prediction
errors) on the test set. 

At this point you should put on your critical / skeptical 
hat and wonder if this did not happen *by chance*, i.e.
if this may be just 
an artifact of the specific training/test partition
we used. The following simple experiment shows that
this is not the case. It would be a **very good exercise** 
for you to repeat it many times (100, say) to verify
my claim.

First, read the whole data and create a new
training / test random split.
```{r cvexperiment1}
# repeat with different partitions
x <- read.csv('rutgers-lib-30861_CSV-1.csv')
set.seed(456)
ii <- sample(rep(1:4, each=15))
x.tr <- x[ii != 2, ]
x.te <- x[ii == 2, ]
```
In the above code chunk, I used `x.tr` to denote the 
training set and `x.te` for the test set. 
Now, fit the full and reduced models 
on this new training set:
```{r cvexperiment2}
full <- lm(MORT ~ . , data=x.tr)
reduced <- lm(MORT ~ POOR + HC + NOX + HOUS + NONW, data=x.tr)
```
Finally, estimate the mean squared prediction
error of these models with their squared prediction
error on the test set:
```{r cvexperiment3}
x.te$pr.full <- predict(full, newdata=x.te)
x.te$pr.reduced <- predict(reduced, newdata=x.te)
with(x.te, mean( (MORT - pr.full)^2 ))
with(x.te, mean( (MORT - pr.reduced)^2 ))
```
Note that the estimated mean squared prediction error
of the reduced model is again considerably smaller
than that of the full model (even though the latter always fits the 
training set better than the reduced one).
