---
title: "STAT406 - Lecture 18 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #, cache=TRUE) #, eval = FALSE)
```

## LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

Preliminary lecture slides will be here.

## What is Adaboost doing, *really*? 

We have seen in class that Adaboost can be thought of as 
fitting an *additive model* in a stepwise (greedy) way,
using an exponential loss. 
It is then easy to prove that Adaboost.M1 
is computing an approximation to the *optimal classifier*
G( x ) = log[ P( Y = 1 | X = x ) / P( Y = -1 | X = x ) ] / 2.
More specifically, Adaboost.M1 is fitting an 
additive model to that function. In other words, it is
attempting to find functions $f_1$, $f_2$, ..., $f_N$ such that 
$G(x) = \sum_i f_i( x^{(i)} )$, where $x^{(i)}$ is a sub-vector
of $x$ (i.e. the function $f_i$ only depends on *some* of the
available features, typically a few of them: 1 or 2, say). Note
that each $f_i$ generally depends on a different subset of 
features than the other $f_j$'s. 

Knowing what function the boosting algorithm is approximating (albeit in 
a greedy and suboptimal way), allows us to   
understand when the algorithm is expected to work well,
and also when it may not work well. 
In particular, it provides one way to choose the complexity of the 
*weak lerners* used to construct the ensemble. For an example
you can refer to the corresponding lab activity. 

### A more challenging example, the `email spam` data

The email spam data set is a relatively classic data set 
containing 57 features (potentially explanatory variables) 
measured on 4601 email messages. The goal is to predict
whether an email is *spam* or not. The 57 features are 
a mix of continuous and discrete variables. More information
can be found at
[https://archive.ics.uci.edu/ml/datasets/spambase](https://archive.ics.uci.edu/ml/datasets/spambase).

We first load the data and randomly separate it into a training and
a test set. A more thorough analysis would be to use 
*full* K-fold cross-validation, but given the computational
complexity, I decided to leave the rest of this 
3-fold CV exercise to the reader. 
```{r spam.1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
data(spam, package='ElemStatLearn')
n <- nrow(spam)
set.seed(987)
ii <- sample(n, floor(n/3))
spam.te <- spam[ii, ]
spam.tr <- spam[-ii, ]
```
We now use Adaboost with 500 iterations, using *stumps* as our
weak learners / classifiers, and check the performance on
the test set:
```{r spam.2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(adabag)
onesplit <- rpart.control(cp=-1, maxdepth=1, minsplit=0, xval=0)
bo1 <- boosting(spam ~ ., data=spam.tr, boos=FALSE, mfinal=500, control=onesplit)
pr1 <- predict(bo1, newdata=spam.te)
table(spam.te$spam, pr1$class) # (pr1$confusion)
```
The classification error rate on the test set is rather high (`r pr1$error`). We now
compare it with that of a Random Forest:

```{r spam.3, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(randomForest)
set.seed(123) 
a <- randomForest(spam ~ . , data=spam.tr) # , ntree=500)
```
We look at the output
```{r spam.3.rf, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
a
```
Note that the OOB estimate of the classification error rate 
is `r a$err.rate[a$ntree, "OOB"]`. 
The number of trees used seems to be appropriate in terms
of the stability of the OOB error rate estimate:
```{r spam.plot.rf, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(a)
```

Now use the test set to estimate the error rate of the Random Forest 
(for a fair comparison with the one computed with boosting) and obtain
```{r spam.error.rate.rf, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
pr.rf <- predict(a, newdata=spam.te, type='response')
table(spam.te$spam, pr.rf)
```
The performance of Random Forests on this test set is better than that of 
boosting (recall that the estimated classification error rate 
for 1-split trees-based Adaboost was 
`r pr1$error`, while for the Random Forest is `r mean( spam.te$spam != pr.rf )` 
on the test set and `r a$err.rate[a$ntree, "OOB"]` using OOB). 

Is there *any room for improvement* for Adaboost? 
As we discussed in class, depending on the interactions that may be 
present in the *true classification function*, we might be able to 
improve our boosting classifier by slightly increasing the complexity
of our base ensemble members. Here we try to use 3-split classification
trees, instead of the 1-split ones used above:
```{r spam.4, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
threesplits <- rpart.control(cp=-1, maxdepth=3, minsplit=0, xval=0)
bo3 <- boosting(spam ~ ., data=spam.tr, boos=FALSE, mfinal=500, control=threesplits)
pr3 <- predict(bo3, newdata=spam.te)
(pr3$confusion)
```
The number of element on the boosting ensemble appears to be
appropriate:
```{r spam.5, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(errorevol(bo3, newdata=spam.te))
```

There is, in fact, a noticeable improvement in performance on this
test set. The estimated classification error rate of AdaBoost using
3-split trees on this test set is
`r pr3$error`. Recall that the estimated classification error rate
for the Random Forest was `r mean( spam.te$spam != pr.rf )`
(or `r a$err.rate[a$ntree, "OOB"]` using OOB). 

As mentioned above you are strongly encouraged to finish this analysis
by doing a complete K-fold CV analysis in order to compare boosting with random 
forests on these data. 

## Gradient boosting

Discussed in class.

## Neural Networks

Discussed in class.

#### An example with a simple neural network

This example using the ISOLET data illustrates the use of simple
neural networks, and also highlights some issues of which it may 
be important to be aware. 

#### Additional resources for discussion (refer to the lecture for context)

* [https://arxiv.org/abs/1412.6572](https://arxiv.org/abs/1412.6572)
* [https://arxiv.org/abs/1312.6199](https://arxiv.org/abs/1312.6199)
* [https://www.axios.com/ai-pioneer-advocates-starting-over-2485537027.html](https://www.axios.com/ai-pioneer-advocates-starting-over-2485537027.html)
* [https://medium.com/intuitionmachine/the-deeply-suspicious-nature-of-backpropagation-9bed5e2b085e](https://medium.com/intuitionmachine/the-deeply-suspicious-nature-of-backpropagation-9bed5e2b085e)


