---
title: "STAT406 - Lecture 19 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #, eval = FALSE)
```

## LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

Preliminary lecture slides are [here](STAT406-18-lecture-19-preliminary.pdf).

# Unsupervised learning

Unsupervised learning methods differ from the 
supervised ones we have studied so far in that 
there is no response variable. The objective is
not related to prediction but rather
to the identification of different possible structures that may be
present in the data. For example, one may be
interested in determining whether the observations 
are "grouped" in some way (clustering), or if the data can be
efficiently represented using fewer variables or features 
(dimension reduction). 

Many of these methods do not rely on any probabilistic
model, and thus there may not be a clear *target* to 
be estimated or approximated. As a consequence,
the conclusions that can be reached from this type
of analyses are often of an exploratory nature. 

## Clustering

A large class of unsupervised learning methods is
collectively called *clustering*. The objective 
can be described as identifying different groups 
of observations that are closer to each other ("clustered") than to those of 
the other groups. The data consist of *n* observations $X_1$, $X_2$, ..., 
$X_n$, each with *p* features. In general, the number of groups is
unknown and needs to be determined from the data.

In this course we will discuss both model-free
and model-based clustering methods. In the first
group we will present K-means (and other 
related methods), and hierarchical (agglomerative) clustering 
methods. Model-based clustering is based on the assumption that the data 
is a random sample, and that the distribution of
the vector of features **X** is a combination of 
different distributions (technically: a *mixture model*). 
In the latter case, for observations belonging to a
group, the vector **X** is assumed 
to have a distribution in a specific (generally parametric) family. 
Some of these model-based methods treat the group labels as 
missing (unobserved) responses, and rely on the assumed model to 
infer those missing labels. 

### K-means, K-means++, K-medoids

Probably the most intuitive and easier to explain
unsupervised clustering algorithm is K-means (and
its variants *K-means++* and K-medoids, a.k.a. *pam*, 
partition around medoids). The specifics of the K-means 
algorithm were discussed in class. Here we will illustrate its use
on a few examples. 

#### UN votes example. 

These data contain the historical voting patterns
of United Nations members. More details can be
found here: 

> Voeten, Erik; Strezhnev, Anton; Bailey, Michael, 2009,
> "United Nations General Assembly Voting Data",
> http://hdl.handle.net/1902.1/12379, Harvard Dataverse, V11

The UN was founded in 1946 and it contains 193 member states. 
The data include only "important" votes, as classified
by the U.S. State Department. The votes for each country 
were coded as follows: Yes (1), Abstain (2), No (3),
Absent (8), Not a Member (9). There were 
368 important votes, and 77 countries
voted in at least 95% of these. We focus on these
UN members. Our goal is to explore whether
voting patterns reflect political
alignments, and also whether countries vote along known
political blocks. Our data consists of 77 observations
with 368 variables each. More information on these data can be found 
[here](https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/12379).

The dataset is organized by vote (resolution), one per row, 
and its columns contain the corresponding vote of each country
(one country per column).
We first read the data, and limit ourselves to resolutions where
every country voted without missing votes:
```{r unvotes, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
X <- read.table(file='unvotes.csv', sep=',', row.names=1, header=TRUE)
X2 <- X[complete.cases(X),]
```
We now compute a K-means partition using the function `kmeans`
with *K = 5*, and look at the resulting groups:
```{r kmeans.un1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
set.seed(123)
b <- kmeans(t(X2), centers=5, iter.max=20, nstart=1)
table(b$cluster)
```
If we run `kmeans` again, we might get a different partition:
```{r kmeans.un2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
b <- kmeans(t(X2), centers=5, iter.max=20, nstart=1)
table(b$cluster)
```
It is better to consider a large number of random starts and
take the **best** found solution (what does *best* mean in this
context? in other words, how does the algorithm decide which one is
the solution it should return?) 
```{r kmeans.un3, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
# Take the best solution out of 1000 random starts
b <- kmeans(t(X2), centers=5, iter.max=20, nstart=1000)
split(colnames(X2), b$cluster)
```
It may be better to look at the groups on a map:

```{r kmeans.map1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
library(rworldmap)
library(countrycode)
these <- countrycode(colnames(X2), 'country.name', 'iso3c')
malDF <- data.frame(country = these, cluster = b$cluster)
# malDF is a data.frame with the ISO3 country names plus a variable to
# merge to the map data

# This line will join your malDF data.frame to the country map data
malMap <- joinCountryData2Map(malDF, joinCode = "ISO3",
                              nameJoinColumn = "country")
# colors()[grep('blue', colors())]
# fill the space on the graphical device
par(mai=c(0,0,0,0),xaxs="i",yaxs="i")
mapCountryData(malMap, nameColumnToPlot="cluster", catMethod = "categorical",
               missingCountryCol = 'white', addLegend=FALSE, mapTitle="",
               colourPalette=c('darkgreen', 'hotpink', 'tomato', 'blueviolet', 'yellow'),
               oceanCol='dodgerblue')
```

We can compare this partition with the one we obtain using PAM (K-medoids), 
which is implemented in the function `pam` of the package `cluster`. Recall
from the discussion in class that `pam` does not need to manipulate the
actual observations, only its pairwise distances (or dissimilarities). In
this case we use Euclidean distances, but it may be interesting to 
explore other distances, particulary in light of the categorical 
nature of the data. Furthermore, to obtain clusters that may
be easier to interpret we use `K = 3`:
```{r un.pam, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
library(cluster)
# Use Euclidean distances
d <- dist(t(X))
# what happens with missing values?
set.seed(123)
a <- pam(d, k=3)
```
Compare the resulting groups with those of *K-means*:
```{r un.pam2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
b <- kmeans(t(X2), centers=3, iter.max=20, nstart=1000)
table(a$clustering)
table(b$cluster)
```
An better visualization is done using the map. interesting 
We plot the 3 groups found by `pam` on the map, followed by those
found by K-means:

```{r un.pam.map, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
these <- countrycode(colnames(X), 'country.name', 'iso3c')
malDF <- data.frame(country = these, cluster = a$clustering)
malMap <- joinCountryData2Map(malDF, joinCode = "ISO3", nameJoinColumn = "country")
par(mai=c(0,0,0,0),xaxs="i",yaxs="i")
mapCountryData(malMap, nameColumnToPlot="cluster", catMethod = "categorical",
               missingCountryCol = 'white', addLegend=FALSE, mapTitle="",
               colourPalette=c('darkgreen', 'hotpink', 'tomato', 'blueviolet', 'yellow'), oceanCol='dodgerblue')
these <- countrycode(colnames(X2), 'country.name', 'iso3c')
malDF <- data.frame(country = these, cluster = b$cluster)
malMap <- joinCountryData2Map(malDF, joinCode = "ISO3", nameJoinColumn = "country")
par(mai=c(0,0,0,0),xaxs="i",yaxs="i")
mapCountryData(malMap, nameColumnToPlot="cluster", catMethod = "categorical",
               missingCountryCol = 'white', addLegend=FALSE, mapTitle="",
               colourPalette=c('yellow', 'tomato', 'blueviolet'),
               oceanCol='dodgerblue')
```

What if we use the L_1 norm instead?

```{r un.pam.inf, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
d <- dist(t(X), method='manhattan')
set.seed(123)
a <- pam(d, k=3)
these <- countrycode(colnames(X), 'country.name', 'iso3c')
malDF <- data.frame(country = these, cluster = a$clustering)
malMap <- joinCountryData2Map(malDF, joinCode = "ISO3", nameJoinColumn = "country")
par(mai=c(0,0,0,0),xaxs="i",yaxs="i")
mapCountryData(malMap, nameColumnToPlot="cluster", catMethod = "categorical",
               missingCountryCol = 'white', addLegend=FALSE, mapTitle="",
               colourPalette=c('darkgreen', 'hotpink', 'tomato', 'blueviolet', 'yellow'),
               oceanCol='dodgerblue')
```

As mentioned before, since the data set does not include a *true label*, the 
comparison between the different results is somewhat subjective, and it often
relies on the knowledge of the subject matter experts. In our example above, this
would mean asking the opinion of a political scientist as to whether these
groupings correspond to known international political blocks or alignments.  

#### Breweries

In this example beer drinkers were asked to rate 9 breweries one 26
attributes, e.g. whether this brewery has a rich tradition; or
whether it makes very good pilsner beer, etc. For each of these
questions, the *judges* reported a score on a 6-point scale
ranging from 1: "not true at all" to 6: "very true". The data
are in the file `breweries.dat`:
```{r brew, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
x <- read.table('breweries.dat', header=FALSE)
x <- t(x)
```

For illustration purposes we use the $L_1$ distance and the
PAM clustering method.
```{r brew2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
d <- dist(x, method='manhattan')
set.seed(123)
a <- pam(d, k=3)
table(a$clustering)
```
To visualize the strength of these cluster partition we use
the silhouette plot discussed in class:
```{r brew.plot, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(a)
```

Since other distances may produce different partitions, an interesting
exercise would be to compare the above clusters with those found using
the Euclidean or  $L_\infy$ norms, for example. 

#### Cancer example

This data contains gene expression levels for 
6830 genes (rows) for 64 cell samples (columns). 
More information can be found here:
[http://genome-www.stanford.edu/nci60/](http://genome-www.stanford.edu/nci60/). 
The data are included in the `ElemStatLearn` package, and also
available on-line: 
[https://web.stanford.edu/~hastie/ElemStatLearn/](https://web.stanford.edu/~hastie/ElemStatLearn/).

We will use K-means to identify 8 possible clusters among the 64 cell samples. 
As discussed in class this exercise can (perhaps more interestingly) be formulated 
in terms of *feature selection*. 

```{r nci, fig.width=6, fig.height=6, message=FALSE, warning=FALSE, tidy=TRUE}
# nci.data <- read.table('http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/nci.data', header=FALSE)
data(nci, package='ElemStatLearn')
ncit <- t(nci)
set.seed(31)
a <- kmeans(ncit, centers=8, iter.max=5000, nstart=100)
table(a$cluster)
split(colnames(nci), a$cluster)
```

