---
title: "STAT406 - Lecture 21 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=6,  
                      warning=FALSE, message=FALSE, tidy=TRUE) #, eval = FALSE)
```

## LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

- Lecture slides are [here](STAT406-19-lecture-21.pdf).
- Lecture activity sheet is [here](STAT406-19-lecture-21-activity.pdf).

## Hierarchical clustering.

Hierarchical clustering refers to a class of algorithms that work in a 
different way from the ones we have seen so far. Both k-means and model-based
clustering try to find a pre-specified number of clusters *simultaneously*. 
Hierarchical methods are *agglomerative*--they start with *n* clusters 
(one singleton cluster for each observation in the data set), and form a
hierarchical 
sequence of clusters of sizes *n-1*, *n-2*, ..., *3*, *2*, and a final "cluster"
composed of all the observations. The user then needs to decide where to 
*cut* the sequence, in other words, how many clusters to identify. 
Algorithms in this class are also called
agglomerative, for obvious reasons. 

The general algorithm can be described as follows:

1. Set *K = n* (the number of observations in the data), and Start with *n* clusters;
2. While *K > 1*:
    a. Merge 2 clusters to form *K-1* clusters;
    b. Set `K = K - 1` (i.e. decrease *K* by one).

The different versions (*flavours*) of this method are obtained 
by varying the criteria to decide which 2 clusters to merge at each 
run of step 2(i) above, which depends on how we measure the distance
(or dissimilarity)  between clusters.

There are a few different tools to decide how many clusters may be present
in the data following a hierarchical clustering algorithm. The most 
commonly used is a graphical representation of the sequence of 
clusters, called a *dendogram*. 

Please refer to your class notes for details on the different
merging criteria (i.e. deciding which clusters to combine
at each step)
and the interpretation of a dendogram. Below
we will illustrate the use of these algorithms on a few
examples. 

#### Breweries example

Beer drinkers were asked to rate 9 breweries on 26
attributes. The attributes were, e.g., Brewery has rich tradition; or
Brewery makes very good Pils beer. Relative to each attribute, the
informant had to assign each brewery a score on a 6-point scale
ranging from 1=not true at all to 6=very true.
We read the data, and use the function `dist` 
to compute the pairwise *L_1* distances between 
the 9 breweries. Note that the data are available
columnwise ($p \times x$) so we first transpose
it before we compute the distances. We also
change the misleading column names assigned
by `read.table`, which are not features but
rather observation numbers:
```{r breweries}
x <- read.table('../Lecture19/breweries.dat', header=FALSE)
colnames(x) <- paste0('Brew-', 1:ncol(x))
x <- t(x)
d <- dist(x, method='manhattan')
```

One implementation of hierarchical clustering methods in `R` is
in the function `hclust` in package `cluster`. 
We first use Ward's information criterion (corrected to 
appropriately use squared distances). The  `plot` method for objects
of class `hclust` produces the associated dendogram. The function `rect.hclust` 
computes the height at which one shuld *cut* the dendogram to
obtain a desired number *k* of clusters. Below we show 
the result for *K = 3* clusters:

```{r breweries.2}
# hierarchical
library(cluster)
# show the dendogram
plot(cl <- hclust(d, method='ward.D2'), main='', xlab='', sub='', hang=-1)
# identify 3 clusters
rect.hclust(cl, k=3, border='red')
```

Now we repeat the analysis but using Euclidean distances and *single linkage*, 
and show *K = 3* clusters:

```{r breweries2}
br.dis <- dist(x) # L2
br.hc <- hclust(br.dis, method='single')
plot(br.hc)
br.hc.3 <- rect.hclust(br.hc, k=3)
```

Note how these 3 clusters are somewhat different from the 
ones found before. However, the *(V1, V4, V7)* cluster
is present in both partitions, and also the triplet 
*(V3, V6, V8)* stays together as well. 
It is interesting to compare these clusters with those
found by K-means (see previous notes), in particular,
these dendograms resemble 
the information on the silhouette plots to some extent. 


#### Languages example

The details of this example were discussed in class. Here we 
present the results of three commonly used merging 
criteria:
*single* linkage, *complete* linkage, *average* linkage, and *Ward's* 
criterion. As usual, we start by reading 
the data, which in this case 
are the specific dissimilarities between languages
discussed in class, and we arrange them in a
matrix that can be used by `hclust`:
```{r languages}
dd <- read.table('languages.dat',  header=FALSE)
names(dd) <- c('E', 'N', 'Da', 'Du', 'G', 'Fr', 'S', 'I', 'P', 'H', 'Fi')
dd <- ( dd + t(dd) / 2)
d <- as.dist(dd)
```
Now we compute a hierarchical clustering sequence using **single linkage**,
plot the corresponding dendogram and identify 4 clusters:

```{r langs2}
plot(cl <- hclust(d, method='single'), main='', xlab='', sub='', hang=-1)
rect.hclust(cl, k=4, border='red')
```

Compare the above with the results obtained with **complete linkage**:

```{r langs3}
plot(cl <- hclust(d, method='complete'), main='', xlab='', sub='', hang=-1)
rect.hclust(cl, k=4, border='red')
```

With **average linkage** we obtain:

```{r langs3.1}
plot(cl <- hclust(d, method='average'), main='', xlab='', sub='', hang=-1)
rect.hclust(cl, k=4, border='red')
```


And finally, using **Ward's criterion** results in the following
dendogram and 4 clusters:

```{r langs4}
plot(cl <- hclust(d, method='ward.D2'), main='', xlab='', sub='', hang=-1)
rect.hclust(cl, k=4, border='red')
```

<!-- # ```{r languages2} -->
<!-- # # read the pairwise dissimilarities -->
<!-- # # (there's no data!) -->
<!-- # a.la <- read.table('languages.dat', header=FALSE) -->
<!-- #  -->
<!-- # # since only the lower triangular matrix is available -->
<!-- # # we need to copy it on the upper half -->
<!-- # a.la <- a.la + t(a.la) -->
<!-- #  -->
<!-- # # create a vector of language names, to be used later -->
<!-- # la.nms <- c('E', 'N', 'Da', 'Du', 'G', 'Fr', 'S', 'I', -->
<!-- # 'P', 'H', 'Fi') -->
<!-- #  -->
<!-- # # compute hierarchical clustering using single linkage -->
<!-- # la.hc <- hclust(as.dist(a.la), method='single') -->
<!-- #  -->
<!-- # # show the dendogram, use labels in object la.nms -->
<!-- # plot(la.hc, labels=la.nms) -->
<!-- #  -->
<!-- # # compute hierarchical clustering using complete linkage -->
<!-- # la.hc <- hclust(as.dist(a.la), method='complete') -->
<!-- #  -->
<!-- # # show the dendogram, use labels in object la.nms -->
<!-- # plot(la.hc, labels=la.nms) -->
<!-- # ``` -->
<!-- #  -->

#### Cancer example

Here we revisit the Cancer example discussed before. We use Euclidean
distances and Ward's information criterion. Below we show the 
clusters identified when we stop the algorithm at *K = 8*, which
based on the dendogram seems to be a reasonable choice:

```{r cancer, fig.width=9}
data(nci, package='ElemStatLearn')
nci.dis <- dist(t(nci), method='euclidean')
plot(nci.hc.w <- hclust(nci.dis, method='ward.D2'), main='', 
     xlab='', sub='', hang=-1, labels=rownames(nci) )
rect.hclust(nci.hc.w, k=8, border='red')
```

For completeness, below we show the results obtained with the other linkage criteria, 
including Ward's:

```{r cancer2, fig.width=9}
nci.hc.s <- hclust(nci.dis, method='single')
nci.hc.c <- hclust(nci.dis, method='complete')
nci.hc.a <- hclust(nci.dis, method='average')

# plot them
plot(nci.hc.s, labels=colnames(nci), cex=.5)

plot(nci.hc.c, labels=colnames(nci), cex=.5)

plot(nci.hc.a, labels=colnames(nci), cex=.5)
```

Note that with these 3 other criteria no clear structure
seems apparent in the data. 


<!-- #### UN Votes -->

<!-- ```{r unvotes} -->
<!-- X <- read.table(file='../Lecture20/unvotes.csv', sep=',', row.names=1, header=TRUE) -->
<!-- ``` -->



#### Nations example

This is a smaller Political Science dataset. Twelve countries
were assessed on their perceived "likeness" by Political Science 
students. Note that (as in the Languages example
above) in this example we do not have raw observations (features),
we only have access to the already determined parwise dissimilarities. 
Below we show the results of using hierarchical clustering with complete and average 
linkage merging criteria, which produce identical clusters. You
are encouraged to investigate what can be found with other 
merging criteria. 

```{r nations}
# read the pairwise dissimilarities
a2 <- read.table('nations2.dat', header=FALSE)

# since only the lower triangular matrix is available
# we need to copy it on the upper half
a2 <- a2 + t(a2)

# create a vector of country names, to be used later
nams2 <- c('BEL', 'BRA', 'CHI', 'CUB', 'EGY', 'FRA',
'IND', 'ISR', 'USA', 'USS', 'YUG', 'ZAI')

# compute hierarchical clustering using complete linkage
na.hc <- hclust(as.dist(a2), method='complete')
plot(na.hc, labels=nams2)

# compute hierarchical clustering using average linkage
na.hc <- hclust(as.dist(a2), method='average')
plot(na.hc, labels=nams2)
```


#### UN Votes

We revisit here the UN votes example (see Lecture 19). Using Euclidean
distances and Ward's criterion we obtain the following 3 clusters:

```{r unvotes} 
X <- read.table(file='../Lecture19/unvotes.csv', sep=',', row.names=1, header=TRUE) 
un.dis <- dist(t(X), method='euclidean')
un.hc <- hclust(un.dis, method='ward.D2')
plot(un.hc, cex=.5)
un.hc.3 <- rect.hclust(un.hc, k=3)
lapply(un.hc.3, names)
``` 

If we repeat the same exercise but using $L_1$ distances we obtain 
different clusters. 

```{r unvotes.l1} 
un.dis.l1 <- dist(t(X), method='manhattan')
un.hc.l1 <- hclust(un.dis.l1, method='ward.D2')
plot(un.hc.l1, cex=.5)
un.hc.l1.3 <- rect.hclust(un.hc.l1, k=3)
lapply(un.hc.l1.3, names)
```

It is easier to compare these 2 sets of clusters
if we show them on a map. We first find the 
cluster labels corresponding to 3 clusters 
using Euclidean and $L_1$ distances:
```{r unvotes.labs}
labs <- cutree(un.hc, k=3)
labs.l1 <- cutree(un.hc.l1, k=3)
```
We can now use these labels to color a map, as
we did previously. For the Euclidean distances
we obtain:

```{r unvotes.maps.1, results='hide'}
library(rworldmap)
library(countrycode)
these <- countrycode(colnames(X), 'country.name', 'iso3c')
malDF <- data.frame(country = these, cluster = labs)
malMap <- joinCountryData2Map(malDF, joinCode = "ISO3", nameJoinColumn = "country")
par(mai=c(0,0,0,0),xaxs="i",yaxs="i")
mapCountryData(malMap, nameColumnToPlot="cluster", catMethod = "categorical",
               missingCountryCol = 'white', addLegend=FALSE, mapTitle="",
               colourPalette=c('darkgreen', 'hotpink', 'tomato', 'blueviolet', 'yellow'),
               oceanCol='dodgerblue')
```

While with the $L_1$ distances we get:

```{r unvotes.maps.2, include=FALSE}
malDF <- data.frame(country = these, cluster = labs.l1)
malMap <- joinCountryData2Map(malDF, joinCode = "ISO3", nameJoinColumn = "country")
```
```{r unvotes.maps.2.2, echo=FALSE}
par(mai=c(0,0,0,0),xaxs="i",yaxs="i")
mapCountryData(malMap, nameColumnToPlot="cluster", catMethod = "categorical",
               missingCountryCol = 'white', addLegend=FALSE, mapTitle="",
               colourPalette=c('darkgreen', 'hotpink', 'tomato', 'blueviolet', 'yellow'),
               oceanCol='dodgerblue')
```

Recall that, as discussed in class, the analyses above may be 
questionable, because these distance measures do not take into
account the actual nature of the available features. 

