---
title: "STAT406 - Lecture 4 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

* A preliminary set of lecture slides are [here](STAT406-18-lecture-4-preliminary.pdf). 
* A reference paper for a formal derivation of AIC: 
Cavanaugh, J.E. (1997). Unifying the derivations for the Akaike 
and corrected Akaike information criteria. 
*Statistics & Probability Letters*, **33**(2), 201-208. 
[DOI: 10.1016/S0167-7152(96)00128-9](https://doi.org/10.1016/S0167-7152(96)00128-9)


## Estimating MSPE with CV when the model was built using the data

Last week we learned that one needs to be careful when using cross-validation (in any of its flavours--leave one out, K-fold, etc.) Misuse of cross-validation is, unfortunately,
not unusual. For [one example](https://doi.org/10.1073/pnas.102102699) see:

> Ambroise, C. and McLachlan, G.J. (2002). Selection bias in gene 
> extraction on the basis of microarray gene-expression data,
> PNAS, 2002, 99 (10), 6562-6566.
> DOI: 10.1073/pnas.102102699

In particular, for every fold one needs to repeat **everything** that was done with the training set (selecting variables, looking at pairwise correlations, AIC values, etc.)

## Correlated covariates

Technological advances in recent decades have resulted in data
being collected in a fundamentally different manner from the way
it was done when most "classical" statistical methods were developed
(early to mid 1900's).
Specifically, it is now not at all uncommon to have data sets with
an abundance of potentially useful explanatory variables 
(for example with more variables than observations). 
Sometimes the investigators are not sure which of the collected variables
can be
expected to be useful or meaningful. 

A consequence of this "wide net" data collection strategy is
that many of the explanatory variables may be correlated with
each other. In what follows we will illustrate some of the
problems that this can cause both when training and interpreting
models, and also with the resulting predictions.

### Variables that were important may suddenly "dissappear"

Consider the air pollution data set we used 
earlier, and the 
**reduced** linear regression model discussed in class:
```{r signif}
x <- read.table('../Lecture1/rutgers-lib-30861_CSV-1.csv', header=TRUE, sep=',')
reduced <- lm(MORT ~ POOR + HC + NOX + HOUS + NONW, data=x)
round( summary(reduced)$coef, 3)
```
Note that all coefficients seem to be significant based on
the individual tests of hypothesis (with `POOR` and
`HOUS` maybe only marginally so). In this sense all 5
explanatory varibles in this model appear to be relevant.

Now, we fit the **full** model, that is, we include
all available explanatory variables in the data set:
```{r signif2}
full <- lm(MORT ~ ., data=x)
round( summary(full)$coef, 3)
```
In the **full** model there 
are many more parameters that need to be estimated, and while two of
them appear to be significantly different from zero (`NONW`
and `PREC`), all the others appear to be redundant.
In particular, note that the p-values for the individual
test of hypotheses for 4 out of the 5
regression coefficients for the variables of the **reduced**
model have now become not significant.
```{r signif3}
round( summary(full)$coef[ names(coef(reduced)), ], 3)
```
In other words, the coeffficients of 
explanatory variables that appeared to 
be relevant in one model may turn
to be "not significant" when other variables
are included. This could pose some challenges 
for interpreting the estimated parameters of the
models.


### Why does this happen?

Recall that the covariance matrix of the least squares estimator involves the
inverse of (X'X), where X' denotes the transpose of the n x p matrix X (that
contains each vector of explanatory variables as a row). It is easy to see
that if two columns of X are linearly dependent, then X'X will be rank deficient.
When two columns of X are "close" to being linearly dependent (e.g. their
linear corrleation is high), then the matrix X'X will be ill-conditioned, and
its inverse will have very large entries. This means that the estimated
standard errors of the least squares estimator will be unduly large, resulting
in non-significant test of hypotheses for each parameter separately, even if
the global test for all of them simultaneously is highly significant.

### Why is this a problem if we are interested in prediction?

Although in many applications one is interested in interpreting the parameters
of the model, even if one is only trying to fit / train a model to do
predictions, highly variable parameter estimators will typically result in
a noticeable loss of prediction accuracy. This can be easily seen from the
bias / variance factorization of the mean squared prediction error (MSPE)
mentioned in class. Hence, better predictions can be obtained if one
uses less-variable parameter (or regression function) estimators.

### What can we do?

A commonly used strategy is to remove some explanatory variables from the
model, leaving only non-redundant covariates. However, this is easier said than
done. You will have seen some strategies in previous Statistics
courses (e.g. stepwise variable selection). 
In coming weeks we will investigate other methods to deal with this problem.


## Comparing models -- General strategy

Suppose we have a set of competing models from which we want to choose the
"best" one. In order to properly define our problem we need the following:

* a list of models to be considered;
* a numerical measure to compare any two models in our list;
* a strategy (algorithm, criterion) to navigate the set of models; and
* a criterion to stop the search. 

For example, in stepwise methods the models under consideration in 
each step are those that differ from the current model only by one
coefficient (variable). The numerical measure used to compare models
could be AIC, or Mallow's Cp, etc. The strategy is to only consider 
submodels with one fewer variable than the current one, and we stop
if either none of these "p-1" submodels is better than the current one, or 
we reach an empty model. 

## Comparing models -- What is AIC?

One intuitively sensible quantity that can be used to compare models is a
distance measuring how "close" the distributions implied by these models are from the actual stochastic process generating the data (here "stochastic process" refers to the random mechanism that generated the observations). In order to do this we need:

a. a distance / metric (or at least a "quasimetric") between models; and 
a. a way of estimating this distance when the "true" model is unknown.

AIC provides an unbiased estimator of the Kullback-Leibler divergence 
between the estimated model and the "true" one. See the lecture slides
for more details. 

## Using stepwise + AIC to select a model

We apply stepwise regression based on AIC to select a linear
regression model for the airpollution data. In `R` we can use
the function `stepAIC` in package `MASS` to perform a stepwise
search, for the synthetic data set discussed in class:
```{R step1}
set.seed(123)
x1 <- rnorm(506)
x2 <- rnorm(506, mean=2, sd=1)
x3 <- rexp(506, rate=1)
x4 <- x2 + rnorm(506, sd=.1)
x5 <- x1 + rnorm(506, sd=.1)
x6 <- x1 - x2 + rnorm(506, sd=.1)
x7 <- x1 + x3 + rnorm(506, sd=.1)
y <- x1*3 + x2/3 + rnorm(506, sd=2.2)

x <- data.frame(y=y, x1=x1, x2=x2,
                x3=x3, x4=x4, x5=x5, x6=x6, x7=x7)

library(MASS)
null <- lm(y ~ 1, data=x)
full <- lm(y ~ ., data=x)
st <- stepAIC(null, scope=list(lower=null, upper=full), trace=FALSE)
```
If you want to see the progression of the search step-by-step, set the
argument `trace=TRUE` in the call to `stepAIC` above. 
The selected model is automatically fit and returned, so that
in the code above `st` is an object of class `lm` containing the
"best" linear regression fit. 
```{r best1}
st
```

We will now compare the mean squared prediction errors of 
the **full** model and that selected with **stepwise**. 
We use 50 runs of 5-fold CV, and obtain
the following:

```{r cv1, fig.width=5, fig.height=5, echo=FALSE}
k <- 5
n <- nrow(x)
ii <- (1:n) %% k + 1
set.seed(123)
N <- 50
mspe.t <- mspe.f <- mspe.st <- rep(0, N)
for(i in 1:N) {
  ii <- sample(ii)
  pr.t <- pr.f <- pr.st <- rep(0, n)
  for(j in 1:k) {
    x0 <- x[ii != j, ]
    null0 <- lm(y~1, data=x0)
    full0 <- lm(y~., data=x0) # needed for stepwise
    true0 <- lm(y~x1 + x2, data=x0)
    step.lm0 <- stepAIC(null0, scope=list(lower=null0, upper=full0), trace=FALSE)
    pr.st[ ii == j ] <- predict(step.lm0, newdata=x[ii==j,])
    pr.f[ ii == j ] <- predict(full0, newdata=x[ii==j,])
    pr.t[ ii == j ] <- predict(true0, newdata=x[ii==j,])
  }
  mspe.st[i] <- mean( (x$y - pr.st)^2 )
  mspe.f[i] <- mean( (x$y - pr.f)^2 )
  mspe.t[i] <- mean( (x$y - pr.t)^2 )
}
boxplot(mspe.st, mspe.f, names=c('Stepwise', 'Full'),
        col=c('gray60', 'hotpink'), ylab='MSPE')
```

Note that since this is a synthetic data set, we can also
estimate the MSPE of the **true** model (could we compute it analytically instead?)
and compare it with that of the **full** and **stepwise** models. 
We obtain:

```{r cv2, fig.width=5, fig.height=5, echo=FALSE}
boxplot(mspe.t, mspe.st, mspe.f, names=c('True', 'Stepwise', 'Full'),
        col=c('tomato', 'gray60', 'hotpink'), ylab='MSPE')
```

### Stepwise applied to the "air pollution" data 

We now use stepwise on the air pollution data to select a model, and
estimate its MSPE using 5-fold CV. We compare the predictions of 
this model with that of the full model. 

```{r stepair}
library(MASS)
airp <- read.table('../lecture1/rutgers-lib-30861_CSV-1.csv', header=TRUE, sep=',')
null <- lm(MORT ~ 1, data=airp)
full <- lm(MORT ~ ., data=airp)
( tmp.st <- stepAIC(full, scope=list(lower=null), trace=FALSE) )
```

```{r mspe.air, fig.width=5, fig.height=5}
k <- 5
n <- nrow(airp)
ii <- (1:n) %% k + 1
set.seed(123)
N <- 50
mspe.f <- mspe.st <- rep(0, N)
for(i in 1:N) {
  ii <- sample(ii)
  pr.f <- pr.st <- rep(0, n)
  for(j in 1:k) {
    x0 <- airp[ii != j, ]
    null0 <- lm(MORT ~ 1, data=x0)
    full0 <- lm(MORT ~ ., data=x0) # needed for stepwise
    step.lm0 <- stepAIC(null0, scope=list(lower=null0, upper=full0), trace=FALSE)
    pr.st[ ii == j ] <- predict(step.lm0, newdata=airp[ii==j,])
    pr.f[ ii == j ] <- predict(full0, newdata=airp[ii==j,])
  }
  mspe.st[i] <- mean( (airp$MORT - pr.st)^2 )
  mspe.f[i] <- mean( (airp$MORT - pr.f)^2 )
}
boxplot(mspe.st, mspe.f, names=c('Stepwise', 'Full'),
        col=c('gray60', 'hotpink'), ylab='MSPE')
```

We can also use the package `leaps` to run a more thorough search
among all possible subsets. We do this with the air pollution data:
```{r leaps1}
library(leaps)
a <- leaps(x=as.matrix(airp[, -16]), y=airp$MORT, int=TRUE, method='Cp', nbest=10)
```
In the call above we asked `leaps` to compute the 10 best models
of each size, according to Mallow's Cp criterion. We can look at
the returned object
```{r leaps.what}
str(a)
```
We now find the best model (based on Mallow's Cp), and 
fit the corresponding model:
```{r leaps.best}
j0 <- which.min(a$Cp)
( m1 <- lm(MORT ~ ., data=airp[, c(a$which[j0,], TRUE)]) )
```
We compare which variables are used in this model with those
used in the model found with stepwise:
```{r leaps.step}
formula(m1)[[3]]
formula(tmp.st)[[3]]
```
It is reasonable to ask whether the model found by `leaps` is 
much better than the one returned by `stepAIC`:
```{r leaps.vs.aic}
extractAIC(m1)
extractAIC(tmp.st)
```
Finally, what is the MSPE of the model found by `leaps`?

```{r mspe.leaps.cv, fig.width=5, fig.height=5}
# proper way
k <- 5
n <- nrow(airp)
ii <- (1:n) %% k + 1
set.seed(123)
N <- 50
mspe.l <- rep(0, N)
for(i in 1:N) {
  ii <- sample(ii)
  pr.l <- rep(0, n)
  for(j in 1:k) {
    x0 <- airp[ii != j, ]
    tmp.leaps <- leaps(x=as.matrix(x0[, -16]), y=as.vector(x0[,16]), int=TRUE, method='Cp', nbest=10)
    j0 <- which.min(tmp.leaps$Cp)
    step.leaps <- lm(MORT ~ ., data=x0[, c(tmp.leaps$which[j0,], TRUE)])
    pr.l[ ii == j ] <- predict(step.leaps, newdata=airp[ii==j,])
  }
  mspe.l[i] <- mean( (airp$MORT - pr.l)^2 )
}
boxplot(mspe.st, mspe.f, mspe.l, names=c('Stepwise', 'Full', 'Leaps'),
        col=c('gray60', 'hotpink', 'steelblue'), ylab='MSPE')
```

Note that a "suboptimal" model (stepwise) seems to be better than
the one found with a "proper" (exhaustive) search, as that returned by
`leaps`. This is intriguing, but we will see the same phenomenon 
occur in different contexts later in the course. 

## An example where one may not need to select variables

In some cases one may not need to select a subset of explanatory
variables, and in fact, doing so may affect negatively the accuracy of
the resulting predictions. In what follows we discuss such an example. 
Consider the credit card data set that contains information
on credit card users. The interest is in predicting the 
balance carried by a client. We first load the data, and to
simplify the presentation here we consider only the numerical
explanatory variables:
```{r credit1}
x <- read.table('Credit.csv', sep=',', header=TRUE, row.names=1)
x <- x[, c(1:6, 11)]
```
There are 6 available covariates, and a stepwise search selects 
a model with 5 of them (discarding `Education`):
```{r credit2, warning=FALSE, message=FALSE}
library(MASS)
null <- lm(Balance ~ 1, data=x)
full <- lm(Balance ~ ., data=x)
(tmp.st <- stepAIC(null, scope=list(lower=null, upper=full), trace=0))
```
It is an easy exercise to check that the MSPE of this
smaller model is in fact worse than the one for the **full** one:

```{r credit3, fig.width=5, fig.height=5, echo=FALSE}
n <- nrow(x)
k <- 5
ii <- (1:n) %% k + 1
set.seed(123)
N <- 100
mspe.st <- mspe.f <- rep(0, N)
for(i in 1:N) {
  ii <- sample(ii)
  pr.f <- pr.st <- rep(0, n)
  for(j in 1:k) {
    null <- lm(Balance ~ 1, data=x[ii != j, ])
    full <- lm(Balance ~ ., data=x[ii != j, ])
    tmp.st <- stepAIC(null, scope=list(lower=null, upper=full), trace=0)
    pr.st[ ii == j ] <- predict(tmp.st, newdata=x[ii==j,])
    pr.f[ ii == j ] <- predict(full, newdata=x[ii==j,])
  }
  mspe.st[i] <- mean( (x$Balance - pr.st)^2 )
  mspe.f[i] <- mean( (x$Balance - pr.f)^2 )
}
boxplot(mspe.st, mspe.f, names=c('Stepwise', 'Full'),
        col=c('tomato', 'springgreen'), cex.axis=1.5,
        cex.lab=1.5, cex.main=2)
mtext(expression(hat(MSPE)), side=2, line=2.5)
```


<!-- ## Shrinkage methods / Ridge regression  -->

<!-- Stepwise methods are highly variable, and thus their predictions may not  -->
<!-- be very accurate (high MSPE).  -->
<!-- A different way to manage correlated explanatory variables (to "reduce" their -->
<!-- presence in the model without removing them) is... -->

<!-- ### Selecting the amount of shrinkage -->




