---
title: "STAT406 - Lecture 13 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

Preliminary lecture slides are [here](STAT406-18-lecture-13-preliminary.pdf).

# Classification

As we discussed in class, what is commonly referred to as *classification* 
can be thought of as prediction, when the responses are classes and we
use a particular loss function (the *0-1 loss* we discussed in class). 
Furthermore, it is easy to show (which we did in class) that the optimal 
classifier (in terms of minimizing the
expected misclassification error) is the one that assigns an observation
to the class with the highest probability of occuring, conditional to the
value of the observed explanatory variables. 

A related discussion about including costs of misclassification 
and the difference between prediction and classification can be found here: [http://www.fharrell.com/post/classification/](http://www.fharrell.com/post/classification/).

Most (if not all) classification methods we will cover in this course can be
simply thought of as different approaches to estimate the conditional probability of
each class, conditional on the value of the explanatory variables. In 
symbols: $P( \mathbf{G} = \mathbf{g} | \mathbf{X} = \mathbf{x}_0)$.
The obvious parallel with what we have done before in this class,
is that many (all?) regression methods we discussed in class are 
different ways of estimating the conditional mean of the response 
variable (conditional on the value of the explanatory
variables). 

As in the regression case, there are different ways to estimate this
optimal predictor / classifier. Some will be model-based, some will
be non-parametric in nature. And some can be considered "restricted" 
non-parametric methods (without relying on a model, but imposing some
other type of constrain on the shape of the classifier). The equivalent
methods for regression with continuous responses are: linear
or non-linear regression as model-based methods; kernel or local regression
as non-parametric methods; and splines or regression trees as 
"constrained" (regularized?) non-parametric methods. 

Below we first discuss model-based methods (Linear and Quadratic
Discriminant Analysis) and later
introduce non-parametric methods (nearest-neighbours and 
classification trees). 

## Linear Discriminant Analysis

Probably the "second easiest to explain" approach (what would "*the* easiest 
to explain" be?) to estimate the above probability is to model the distribution of 
the explanatory variables *within* each class. This will then uniquely determine the 
probabilities we need to estimate, as discussed above and in class. 
In particular, one the simplest models we can use is a normal (Gaussian) one. 
As we saw in class, if we assume that the distribution of the features
for each class is Gaussian with a common covariance matrix across clases, then
it easy to show (*and I strongly suggest that you do it*) that the optimal
classifier (using the 0-1 loss function mentioned above) is a linear
function of the explanatory variables. The coefficients of this linear
function depend on the parameters of the assumed Gaussian distributions. The
latter can be estimated using the training set and plugging them in provides
a natural estimator of the optimal classifier. 

The function `lda` in 
the `MASS` library implements this simple classifier. We illustrate it
here on the rather simple and well-known vaso constriction data, available in the 
`robustbase` package. More details, as usual, can be found on its
help page. The response variable takes two values (represented below
as **blue** and **red**), and there are only two
explanatory variables (which allows us to visualize our methods and results). 

```{r lda1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
data(vaso, package='robustbase')
plot(Volume ~ Rate, data=vaso, pch=19, cex=1.5, 
       col=c('red', 'blue')[Y+1])
```

To train the LDA classifier we use the function `lda` as follows (note the
*model-like* syntax to indicate the response and explanatory variables):

```{r lda1.1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(MASS)
a.lda <- lda(Y ~ Volume + Rate, data=vaso)
```

Now, given any value of the explanatory variables `(Volume, Rate)` we
can use the method `predict` on the object returned by `lda()` to 
estimate the conditional probabilities of **blue** and **red**. 
Next we visualize which regions of the feature space will be predicted to 
contain **blue** points (and then obviously which areas will be 
predicted to correspond to **red** responses). To do this we 
build a relatively fine grid of posible `(Volume, Rate)` values
and plot the corresponding predictions for one class
(i.e. the conditional probabilites for that class as a function of
the explanatory variables):

```{r lda1.2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
# build a grid of points of (Volume, Rate),
# and obtain predictions for each of them
# 40,000 in total
xvol <- seq(0, 4, length=200)
xrat <- seq(0, 4, length=200)
xx <- expand.grid(xvol, xrat)
names(xx) <- c('Volume', 'Rate')

pr.lda <- predict(a.lda, newdata=xx)$posterior[,2]
image(xrat, xvol, matrix(pr.lda, 200, 200), col=terrain.colors(100),
      ylab='Volume', xlab='Rate', main='LDA')
points(Volume ~ Rate, data=vaso, pch=19, cex=1.5,
       col=c('red', 'blue')[Y+1])
```

On the plot above larger numbers are shown with lighther colors
(dark green corresponds to very low values). 

This model-based approach to classification (LDA) is optimal
if the model is correct. The strongest assumption of this model
is, of course, the Gaussian distribution of the vector of 
explanatory variables. The second strongest assumption is
that of equal "shape" (covariance matrix) for the distribution
of the explanatory variables across the different groups
in the population. This latter assumption 
can be relaxed slightly if
we assume instead that the features have a Gaussian distribution within each
class, but that the covariance matrix may be different across 
classes. The corresponding optimal classifier is now a quadratic
function of the predictors (**prove it!**). The function `qda` 
in the `MASS` library implements this classifier, and it
can be used just like `lda` (as usual, refer to its help page for details).

This approach can be used with any number of classes. Any limitations? 

## Logistic regression (Review)

If we model the distribution of the  features within each class using a 
multivariate Gaussian distribution, then it is easy to see that the 
boundaries between classes are linear (functions of the features) (*check!*)
Furthermore, the log of the odds ratio between classes is a linear
function. It is interesting to note that one can start with this last
assumption (instead of the full Gaussian model) and arrive at a
fully parametric model for the conditional distibution of the classes
given the features (see the class slides). The parameters can be
estimated using maximum likelihood. For two classes this is the
logistic regression model, which you may have seen in previous
courses. 

We illustrate this on the `vaso` data as before. Since this is
a 2-class problem, we just need to fit a logistic regression model.
The function `glm` in `R` does it for us, we specify that we 
want to fit such a model using the argument `family=binomial`. 
Once we obtain parameter estimators (in the `glm` object `a` below), 
we use the `predict` method to obtain predicted conditional 
probabilities on the same grid we used before:
```{r logistic1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
a <- glm(Y ~ ., data=vaso, family=binomial)
pr <- predict(a, newdata=xx, type='response')
```

We now plot the data and the *surface* of predicted probabilities for 
blue points (higher probabilites are displayed with lighter colors). 

```{r logistic2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
image(xrat, xvol, matrix(pr, 200, 200), col=terrain.colors(100),
      ylab='Volume', xlab='Rate', main='Logistic')
points(Volume ~ Rate, data=vaso, pch=19, cex=1.5,
       col=c('red', 'blue')[Y+1])
```


