---
title: "STAT406 - Lecture 22 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=6,  
                      warning=FALSE, message=FALSE, tidy=TRUE) #, eval = FALSE)
```


## LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

Preliminary lecture slides are [here](STAT406-18-lecture-21-preliminary.pdf).

## Principal Components Analysis

Although principal components can be easily computed with the spectral
decomposition of the covariance matrix of the data (using the function 
`svd` in `R`, for example), there are a few dedicated implementations
in `R`, among them `prcomp` and `princomp`). The main difference between
these two is which internal function is used to compute eigenvalues and
eigenvectors: `prcomp` uses `svd` and `princomp` uses the less preferred
function `eigen`. Both `princomp` and `prcomp` return the matrix of 
loadings (eigenvectors), the scores (projections of the data on the 
basis of eigenvectors), and other auxiliary objects. They also include
plot and summary methods.

Instead of reviewing those (which can be easily done individually), in
these notes I will reproduce two of the examples used in class (the simple
2-dimensional one used to motivate the topic, and the more interesting
256-dimensional one using the digits data). Finally, I will also 
show that principal components can be computed using an iterative
algorithm (alternate regression), which may be faster than factorizing
the covariance matrix, particularly when one is only interested in a
few principal components and the dimension of the data is large. 


#### Simple 2-dimensional example 

We first read the data for the simple illustration of PC's as best lower 
dimensional approximations.
```{r pca1}
x <- read.table('t8-5.dat', header=FALSE)
```
Note that the data has 5 explanatory variables. Here 
we only use two of them in order to be able to visualize the analysis
more easily:
```{r pca2}
xx <- x[,c(2,5)]
colnames(xx) <- c('Prof degree', 'Median home value')
```
As discussed in class, we standardize the data to avoid a large difference
in scales "hijacking" the principal components:
```{r pca3}
xx <- scale(xx, center=colMeans(xx), scale=TRUE)
```
We now define two auxiliary functions to compute Euclidean
norms and squared Euclidean norms (less general by probably
faster than `R`'s `base::norm`):
```{r norms}
norm2 <- function(a) sum(a^2)
norm <- function(a) sqrt(norm2(a))
```

We start by looking at the data with a scatter plot:

```{r data}
plot(xx, pch=19, col='black', cex=2, xlim=c(-1.5, 4.5), ylim=c(-1.5, 4.5))
abline(h=0, lwd=2, col='grey', lty=2)
abline(v=0, lwd=2, col='grey', lty=2)
```

We now compute projections along the direction of the vector
$\mathbf{v} \propto (1, 0.05)^\top$. Recall from your linear algebra courses that
the orthogonal projection of a point $\mathbf{x}$ on the linear subspace spanned by $\mathbf{v}$
(where $\| \mathbf{v} \| = 1$)
is given by $\pi_{\mathbf{v}} ( \mathbf{x} ) = \langle \mathbf{x}, \mathbf{v} \rangle \, \mathbf{v}$
which can also be written as 
$\pi_{\mathbf{v}} ( \mathbf{x} ) = ( \mathbf{v} \, \mathbf{v}^\top) \mathbf{x}$. 
We first find the coordinates of the orthogonal projects of each observation 
along the subspace generated by $\mathbf{v} = (1, 0.05)^\top$ (these are
the scalars
$\langle \mathbf{x}_i, \mathbf{v} \rangle =  \mathbf{x}_i^\top \mathbf{v}$ for each 
point $\mathbf{x}_i$:
```{r proj1}
a <- c(1, 0.05)
a <- a / norm(a)
# Find the projections (coordinates of the
# observations on this basis of size 1)
prs <- (xx %*% a)
```
We now compute the projections 
$\pi_{\mathbf{v}} ( \mathbf{x}_i ) = \langle \mathbf{x}_i, \mathbf{v} \rangle \, \mathbf{v}$:
```{r proj1.1}
pr <- prs %*% a 
```
and add them to the plot, with a few observations highlighted. The subspace
is shown in red, and the orthogonal projections as solid red dots on that line:
```{r proj1.2}
# Plot the data
plot(xx, pch=19, col='black', cex=2, xlim=c(-1.5, 4.5), ylim=c(-1.5, 4.5))
abline(h=0, lwd=2, col='grey', lty=2)
abline(v=0, lwd=2, col='grey', lty=2)
# Show the subspace on which we are projecting
abline(0, a[2]/a[1], lwd=2, col='red')
# Add the projections of the data on this subspace
points(pr[,1], pr[,2], pch=19, cex=1.5, col='red')
# Highlight a few of them
ind <- c(26, 25, 48, 36)
pr2 <- pr[ind, ]
for(j in 1:length(ind))
lines(c(xx[ind[j],1], pr2[j,1]), c(xx[ind[j],2], pr2[j,2]), 
col='blue', lwd=3.5, lty=2)
```

We repeat the above but projecting on a different direction
$\mathbf{v} \propto (-1, 3)^\top$: 

```{r proj2, echo=FALSE}
a <- c(-1, 3)
a <- a / norm(a)
# Find the projections (coordinates of the
# observations on this basis of size 1)
prs <- (xx %*% a)
# Find the orthogonal projections of each
# observation on this subspace of dimension 1
pr <- prs %*% a 
# Plot the data
plot(xx, pch=19, col='black', cex=2, xlim=c(-1.5, 4.5), ylim=c(-1.5, 4.5))
abline(h=0, lwd=2, col='grey', lty=2)
abline(v=0, lwd=2, col='grey', lty=2)

# Show the subspace on which we are projecting
abline(0, a[2]/a[1], lwd=2, col='red')

# Add the projections of the data on this subspace
points(pr[,1], pr[,2], pch=19, cex=1.5, col='red')

# Highlight a few of them
ind <- c(26, 25, 48, 36)
pr2 <- pr[ind, ]
for(j in 1:length(ind))
lines(c(xx[ind[j],1], pr2[j,1]), c(xx[ind[j],2], pr2[j,2]), 
col='blue', lwd=3.5, lty=2)
```

We saw in class that the direction $\mathbf{v}$ that results 
in orthogonal projections closest to the original data (in the sense of 
minimizing the mean (or sum) of the residuals Euclidean norm squared)
is given by the "first" eigenvector of the covariance
matrix of the data. This is the first principal component. 
Refer to the class slides and discussion for more details
and the definition and properties of the other principal components. 

#### Digits example

In this example we use principal components to explore the zip code 
data. In particular, we focus on images from a single digit (we
use 3, but the reader is strongly encouraged to re-do this analysis
for other digits to explore whether similar conclusions hold for 
them). We load the training data from the `ElemStatLearn` package in R,
and extract the images that correspond to the digit 3. For more
information use `help(zip.train, package='ElemStatLearn')`.
```{r digits1}
data(zip.train, package="ElemStatLearn")
a <- zip.train[ zip.train[, 1] == 3, -1 ] 
```
Define an auxiliary function to compute 
the squared Euclidean distance between two vectors
(recall that we have already defined the function `norm2` above):
```{r aux2}
dist <- function(a, b) norm2(a-b)
```

To display the images we adapt the following function
for plotting a matrix, which was 
originally available at [http://www.phaget4.org/R/image_matrix.html](http://www.phaget4.org/R/image_matrix.html):
```{r plotfunction}
myImagePlot <- function(x) {
  min <- min(x)
  max <- max(x)
  ColorRamp <- grey(seq(1,0,length=256))
  ColorLevels <- seq(min, max, length=length(ColorRamp))
  # Reverse Y axis
  reverse <- nrow(x) : 1
  x <- x[reverse,]
  image(1:ncol(x), 1:nrow(x), t(x), col=ColorRamp, xlab="",
        ylab="", axes=FALSE, zlim=c(min,max))
}
```
Using this function, we plot 9 randomly chosen images from our data set:

```{r digitsplot}
set.seed(31)
sa <- sample(nrow(a), 9)
par(mai = c(1, 1, 1, 1)/5, xaxs = "i", yaxs = "i")
par(mfrow=c(3,3))
for(j in 1:9) myImagePlot(t(matrix(unlist(a[sa[j],]), 16, 16)))
```

<!-- # plot the average 3 -->
<!-- myImagePlot(t(matrix(colMeans(a), 16, 16))) -->


<!-- # Plot the first 3 on the data set -->
<!-- myImagePlot(t(matrix(unlist(a[1,]), 16, 16))) -->

Next, we centre the observations in order to compute the eigenvectors 
and eigenvalues of the covariance matrix more efficiently. In fact,
note that we do not need to even compute the covariance matrix
and can use the SVD of the centred data.
```{r digits3}
ac <- scale(a, center=TRUE, scale=FALSE)
si.svd <- svd(ac)
```
<!-- Using that the eigenvalues of the covariance matrix are the squares of the -->
<!-- singular values of the $n \times p$ data matrix, we  -->
<!-- compute the proportion of the "total variance" that is accounted by  -->
<!-- the first 20 ppal components individually -->
<!-- ```{r digits4} -->
<!-- si.svd$d[1:20]^2 / sum(si.svd$d^2) -->
<!-- ``` -->
<!-- and cumulatively: -->
<!-- ```{r digits4.1} -->
<!-- cumsum(si.svd$d^2)[1:20] / sum(si.svd$d^2) -->
<!-- ``` -->
Using the relationship between the eigenvectors of the covariance matrix 
and the SVD of the $n \times p$ data matrix, we compute the coordinates of the 
centered data on their orthogonal projections along each of the first and 2nd and 3rd 
principal directions (eigenvectors of the covariance matrix). Recall that the data 
are stored as rows of the matrix `a`:
```{r digits.p1}
v1 <- as.vector( ac %*% si.svd$v[,1] )
v2 <- as.vector( ac %*% si.svd$v[,2] )
```
As discussed in class, we identify 5 quantiles of each of these 
coordinates to use as our 2-dimensional grid:
```{r digits.p2}
qv1 <- quantile(v1, c(.05, .25, .5, .75, .95))
qv2 <- quantile(v2, c(.05, .25, .5, .75, .95))
```
We can visualize the grid of these 5 x 5 = 25 points 
over the scatter plot of all the 2-dimensional projections 
of the data (their coordinates on the principal components
basis):

```{r digits.p3}
qv <- expand.grid(qv1, qv2)
plot(v1, v2, pch=19, cex=1, col='grey')
points(qv[,1], qv[,2], pch=19, cex=1.5, col='red')
```

We now find the points in our data set (images) 
with projections closest to each of the 5 x 5 = 25 points in the grid
(note that these distances between points in the principal-subspace,
which is in the 256 dimensional space) can be computed 
in terms of their coordinates on the principal-basis only (which 
are 2-dimensional points):
```{r digits.p4}
vs <- cbind(v1, v2)
cvs <- array(0, dim=dim(qv))
for(j in 1:dim(qv)[1]) cvs[j,] <- vs[ which.min( apply(vs, 1, dist, b=qv[j,]) ), ]
```
We now add these points to our plot (we use color blue for them):

```{r digits.p5}
plot(v1, v2, pch=19, cex=1, col='grey')
points(qv[,1], qv[,2], pch=19, cex=1.5, col='red')
for(j in 1:dim(qv)[1]) points(cvs[j,1], cvs[j,2], pch=19, col='blue')
```

Using these "blue" coordinates, we construct the corresponding points 
in the 256-dimensional space:
```{r digits.p6}
app <- t( si.svd$v[,1:2] %*% t( cvs ) )
```
and identify the images in our data set that are closest to these points
```{r digits.p7}
repre <- matrix(0, dim(qv)[1], dim(app)[2])
for(j in 1:dim(qv)[1]) repre[j,] <- ac[ which.min( apply(ac, 1, dist, b=app[j,]) ), ]
```
These are the actual images that are closest to the points in the array `app` 
above. Now add the column means and display these 25 images according to the points
they represent in the red grid:

```{r digits.p8}
repre <- scale(repre, center = -colMeans(a), scale=FALSE)
par(mai = c(1, 1, 1, 1)/5, xaxs = "i", yaxs = "i")
par(mfrow=c(5,5))
for(j in 1:dim(repre)[1]) {
myImagePlot(t(matrix(unlist(repre[j,]), 16, 16)))
}
# par(mfrow=c(1,1))
```

Note how these images change when we "traverse" the 256-dimensional space
along each of these 2 principal directions. 

<!-- Repeat with 3 ppal components now!: -->

<!-- ```{r digits.3pca} -->
<!-- qv1 <- quantile(v1, c(.05, .25, .5, .75, .95)) -->
<!-- qv2 <- quantile(v2, c(.05, .25, .5, .75, .95)) -->
<!-- qv3 <- quantile(v3, c(.05, .5, .95)) -->
<!-- qv <- expand.grid(qv1, qv2, qv3) -->

<!-- vs <- cbind(v1, v2, v3) -->
<!-- # closest v's -->
<!-- cvs <- array(0, dim=dim(qv)) -->
<!-- for(j in 1:dim(qv)[1]) { -->
<!-- cvs[j,] <- vs[ which.min( apply(vs, 1, dist, b=qv[j,]) ), ] -->
<!-- } -->

<!-- app <- t( si.svd$v[,1:3] %*% t( cvs ) ) -->


<!-- repre <- matrix(0, dim(qv)[1], dim(app)[2]) -->

<!-- for(j in 1:dim(qv)[1]) { -->
<!-- repre[j,] <- ac[ which.min( apply(ac, 1, dist, b=app[j,]) ), ] -->
<!-- } -->

<!-- repre <- scale(repre, center = -colMeans(a), scale=FALSE) -->
<!-- par(mai = c(1, 1, 1, 1)/5, xaxs = "i", yaxs = "i") -->
<!-- par(mfrow=c(5,5)) -->
<!-- for(j in 1:dim(repre)[1]) { -->
<!-- myImagePlot(t(matrix(unlist(repre[j,]), 16, 16))) -->
<!-- } -->
<!-- ``` -->

#### Alternating regression to compute principal components

For details see [this document](pca-alternating-regression.pdf).

A function implementing this method to compute the first
principal component is:
```{r alt1}
  alter.pca.k1 <- function(x, max.it = 500, eps=1e-10) {
       n2 <- function(a) sum(a^2)
       p <- dim(x)[2]
       x <- scale(x, scale=FALSE)
       it <- 0
       old.a <- c(1, rep(0, p-1))
       err <- 10*eps
       while( ((it <- it + 1) < max.it) & (abs(err) > eps) ) {
              b <- as.vector( x %*% old.a ) / n2(old.a)
              a <- as.vector( t(x) %*% b ) / n2(b)
              a <- a / sqrt(n2(a))
              err <- sqrt(n2(a - old.a))
              old.a <- a
       }
       conv <- (it < max.it)
       return(list(a=a, b=b, conv=conv))
  }
```
We use it on the digits data above to compute the 
first principal component:
```{r alt2}
tmp <- alter.pca.k1(ac)$a 
```
and compare it with the one given by `svd` (note that the
sign of the eigenvectors is arbitrary, so we adjust 
these vectors in order to have first elements with the 
same sign):
```{r alt3}
tmp2 <- svd(ac)$v[, 1] 
tmp <- tmp * sign(tmp2[1]*tmp[1])
summary(abs(tmp-tmp2))
```
Note that both eigenvectors are essentially identical, and that the 
alternating regression method was approximately 3 times faster than
a full SVD decomposition of the covariance matrix. 

To further illustrate the potential gain in speed for larger dimensions, 
consider the following synthetic data set with n = 2000 observation
and p = 1000, and compare the timing and the results:
```{r alt4}
n <- 2000
p <- 1000
x <- matrix(rt(n*p, df=2), n, p)
system.time( tmp <- alter.pca.k1(x) )
a1 <- tmp$a
system.time( e1 <- svd(cov(x))$u[,1] )
a1 <- a1 * sign(e1[1]*a1[1])
summary( abs(e1 - a1) )
```
